---
title: "Data 607 Project 1"
author: "Jean Jimenez"
date: "2023-09-20"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Project 1

## Overview

Project 1 assignment was to take a text file and to process it and return a csv.

The text file is of chess tournament scores and it is in the format of Chess Ranking System.

In this assignment, I will show how I processed the given data set and exported as csv.

## Work and Thought Process

### Importing, Reading, and Cleaning

The first step of the project was to import the text file into R and read it. I imported the text file using the url (I uploaded it to github). Read lines go line by line. Afterwards, I start cleaning the text. I first eliminate the header rows.

```{r}
#First import text file
#will read line by line

lines= readLines(url("https://raw.githubusercontent.com/sleepysloth12/data607_proj1/main/tournamentinfo.txt"))
lines[1]

#filtering out header

lines=lines[-c(1,2,3,4)]
lines[1]
```

Now that the header is removed, I want to remove the dashed line that divides every player from each other in the text file. The dash line occurs every third row. To remove them, I run it through this if statement. The statement checks to see if the number of rows is divisible by 3. Afterwards, I generate a list of all multiples of 3 until the length of the text files, and I remove those lines (which remove the dashed lines). As you see in the output, Line 3 before the statement is the dashed line, now it is first line of the second player.

```{r}
#making a conditional statement

#If the length of the # of lines is divisible by 3,
#take every third line (dashed) and remove it from our text
#aka generate a all multiples of 3 till len lines

lines[3]

if (length(lines)%%3 == 0){
  multiples_of_3=seq(3, length(lines),by=3)
  lines=lines[-c(multiples_of_3)]
}

lines[3]
```

Now we essentially have two lines for each player. The first line having the player name, points, etc. The second line having State, Pre-rank, etc. I will make two vectors to put each line in them. In other words, the first line of each player will go into one, while the second line will to to two. To do this, I run a for loop that takes every other line (by checking if index is divisible by 2) and puts it in the appropriate vector.

```{r}
#Now we only have player info, with the same info in every other line
#the first line has the names and who is playing against/ win or lose
#the second line has the ranking/ state

#going to separate into two vectors separating two lines 

line_one=c()
line_two=c()
id=1

for(line in lines){
  if(id%%2==0){
    line_two=c(line_two,line)
  }else {
    line_one=c(line_one,line)
  }
  id=id+1
}
line_one[1:5]
line_two[1:5]
```

Now, everything in line_one and everything in line_two has all the same info. The problem now is that it is still in text form. I want it to be in a data frame so that we can do stuff to it. I will now parse out each of the individual sections to get the text between the "\|" character. I do this by using str_split. Then, I convert what I have for each into a matrix, and subsequently a data frame.

```{r}
#Now line one and two are each showing the same info.
#will make each a data frame
# every value separated by  '|'  will get its own column

split_data_one=lapply(line_one, function(x) strsplit(x, "\\|"))
split_data_two=lapply(line_two, function(x) strsplit(x, "\\|"))

#moving from list to matrix [temp] to dataframe

split_data_one_mat=do.call(rbind, lapply(split_data_one, function(x) unlist(x[[1]])))
df_one=as.data.frame(split_data_one_mat, stringsAsFactors = FALSE)
split_data_two_mat=do.call(rbind, lapply(split_data_two, function(x) unlist(x[[1]])))
df_two=as.data.frame(split_data_two_mat, stringsAsFactors = FALSE)

#Adding column names
col_names_1=c('player_id','name','total_pts','rnd_1_comb','rnd_2_comb','rnd_3_comb','rnd_4_comb','rnd_5_comb','rnd_6_comb','rnd_7_comb')
colnames(df_one)=col_names_1
col_names_2=c('state','comb_rank','idk','col_rnd_1','col_rnd_2','col_rnd_3','col_rnd_4','col_rnd_5','col_rnd_6','col_rnd_7')
colnames(df_two)=col_names_2

head(df_one)
head(df_two)
```

### Editing the Data Frame

Now, we have two data frames. We have to modify them a bit. In the first data frame, the round columns show W/L/D as well as the opponent's player ID. For the assignment, we only care about the opponents player id. I used regex to only get the numbers from those columns. Then, I renamed the columns into more appropriate column names since the round columns are no longer combined with other text.

```{r}
#Splitting each combined round x colomn n to get the opponent number
#put in new dataframe

df_one_to_split=c('rnd_1_comb','rnd_2_comb','rnd_3_comb','rnd_4_comb','rnd_5_comb','rnd_6_comb','rnd_7_comb')

#getting only numbeer characters from each column and converting to number

for (col in df_one_to_split){
  df_one[[col]]=as.numeric(gsub("[^0-9]","",df_one[[col]]))
}

head(df_one)

#changing column names to be more accurate
col_names_1=c('player_id','name','total_pts','rnd_1_op','rnd_2_op','rnd_3_op','rnd_4_op','rnd_5_op','rnd_6_op','rnd_7_op')
colnames(df_one)=col_names_1

```

Similarly in the other data set, I had to extract the Pre-rank score of each player. The ranking was located between the 'R:' and the '-\>'. I again used regex to extract this number. Afterwards, I changed column names.

```{r}
#now splitting each combined_rank columnn to get pre rating

library(stringr)

df_two$comb_rank=str_extract(df_two$comb_rank, "(?<=R: )\\d+")
df_two$comb_rank=as.numeric(df_two$comb_rank)
col_names_2=c('state','pre_rank','idk','col_rnd_1','col_rnd_2','col_rnd_3','col_rnd_4','col_rnd_5','col_rnd_6','col_rnd_7')
colnames(df_two)=col_names_2
head(df_two)

```

### Combining and Calculating

I want to combine both data frames into one. First, I removed columns with information that doesn't matter to us. Then I merge both data frames.

```{r}
#Now lets combine both data frames
#first will remove columns we dont care about
df_two$col_rnd_1=NULL
df_two$col_rnd_2=NULL
df_two$col_rnd_3=NULL
df_two$col_rnd_4=NULL
df_two$col_rnd_5=NULL
df_two$col_rnd_6=NULL
df_two$col_rnd_7=NULL
df_two$idk=NULL


#combine the two data frames

chess_df=cbind(df_one,df_two)
chess_df$player_id=as.integer(chess_df$player_id)
head(chess_df)

```

Afterwards, I made a new column to place the average opponent pre rank score. Then I made a for loop to retrieve and calculate this. For each row of the data frame, the loop obtains all of the opponent ids. I did this using regex. Then, I converted all those opponent ids to integers. Afterwards, I took out NA values since some players didn't have opponents for some rounds. Then, I got each of those opponents rank and took the mean. I placed that mean in the new column created.

```{r}
#creating new column named avg op rank
#making a loop that will get all the oposition ranks


chess_df$avg_op_pre_rank = NA

# Loop through each row in chess_df
for (i in 1:nrow(chess_df)) {
  
  # extract op ID for each round
  
  op_ids = unlist(chess_df[i, grep("rnd_\\d+_op", names(chess_df))])
  
  op_ids=as.integer(op_ids)
  
  #take out NA
  
  op_ids=na.omit(op_ids)
  
  # get prerank for each op
  
  op_ranks = chess_df[chess_df$player_id %in% op_ids, "pre_rank"]
  
  #mean op pre rank
  
  chess_df$avg_op_pre_rank[i] = mean(op_ranks, na.rm = TRUE)
}

head(chess_df)
```

### Exporting Data

After I had everything the assignment was asking for, I organized a new data frame with everything that was wanted on the csv file. Then, I wrote the data frame into a new csv file "extracted_chess_info.csv". This new csv file will be placed in the current working directory.

```{r}
#making new dataframe with required format like the assignment states

to_be_exported_list=list(Player_Name=chess_df$name,
                         Player_State=chess_df$state,
                         Total_Points=chess_df$total_pts,
                         Player_Pre_Rating=chess_df$pre_rank,
                         Average_Opponent_Pre_Rating=chess_df$avg_op_pre_rank)
to_be_exported_df=as.data.frame(to_be_exported_list)

head(to_be_exported_df)

#write and export as csv
write.csv(to_be_exported_df,"extracted_chess_info.csv", row.names=FALSE)
```

## Making a Function

Although not required, I felt the need to package all of this into a function. When you do this, you can call it and use it whenever you want without coding again. This is very useful especially when it comes to automation. The following is my function. It is just the same code as above but with the input vars changed. It will return the csv file. I also run an example. You can change the path to your own and run it.

```{r}
chess_to_csv=function(txt_url, export_path_name){
  # chess_to_csv
  #by Jean Jimenez
  #09/20/2023 for Data 607 Proj 1
  
  #Function that takes .txt file of chess tournament notation to .csv to the specifications of the assignment
  
  #INPUTS
  
  #txt_url: url of txt file in chess notation
  
  #export_path_name:  Path and name of csv file to be exported i.e.) "/home/Data607/proj1/chess_test.csv"
  
  
  #OUTPUT
  #CSV file showing player name, player state, total points, pre rating, avg opponent pre rating

  
  #First import text file
  #will read line by line
  
  lines= readLines(txt_url)
  lines
  
  #filtering out header
  
  lines=lines[-c(1,2,3,4)]
  
  
  #making a conditional statement
  
  #If the length of the # of lines is divisible by 3,
  #take every third line (dashed) and remove it from our text
  #aka generate a all multiples of 3 till len lines
  
  
  
  if (length(lines)%%3 == 0){
    multiples_of_3=seq(3, length(lines),by=3)
    lines=lines[-c(multiples_of_3)]
  }
  
  lines
  
  #Now we only have player info, with the same info in every other line
  #the first line has the names and who is playing against/ win or lose
  #the second line has the ranking/ state
  
  #going to separate into two vectors separating two lines 
  
  line_one=c()
  line_two=c()
  id=1
  
  for(line in lines){
    if(id%%2==0){
      line_two=c(line_two,line)
    }else {
      line_one=c(line_one,line)
    }
    id=id+1
  }
  line_one
  line_two
  
  #Now line one and two are each showing the same info.
  #will make each a data frame
  # every value separated by  '|'  will get its own column
  
  split_data_one=lapply(line_one, function(x) strsplit(x, "\\|"))
  split_data_two=lapply(line_two, function(x) strsplit(x, "\\|"))
  
  #moving from list to matrix [temp] to dataframe
  
  split_data_one_mat=do.call(rbind, lapply(split_data_one, function(x) unlist(x[[1]])))
  df_one=as.data.frame(split_data_one_mat, stringsAsFactors = FALSE)
  split_data_two_mat=do.call(rbind, lapply(split_data_two, function(x) unlist(x[[1]])))
  df_two=as.data.frame(split_data_two_mat, stringsAsFactors = FALSE)
  
  #Adding column names
  col_names_1=c('player_id','name','total_pts','rnd_1_comb','rnd_2_comb','rnd_3_comb','rnd_4_comb','rnd_5_comb','rnd_6_comb','rnd_7_comb')
  colnames(df_one)=col_names_1
  col_names_2=c('state','comb_rank','idk','col_rnd_1','col_rnd_2','col_rnd_3','col_rnd_4','col_rnd_5','col_rnd_6','col_rnd_7')
  colnames(df_two)=col_names_2
  
  head(df_one)
  head(df_two)
  
  #Splitting each combined round x colomn n to get the opponent number
  #put in new dataframe
  
  df_one_to_split=c('rnd_1_comb','rnd_2_comb','rnd_3_comb','rnd_4_comb','rnd_5_comb','rnd_6_comb','rnd_7_comb')
  
  #getting only numbeer characters from each column and converting to number
  
  for (col in df_one_to_split){
    df_one[[col]]=as.numeric(gsub("[^0-9]","",df_one[[col]]))
  }
  
  head(df_one)
  
  #changing column names to be more accurate
  col_names_1=c('player_id','name','total_pts','rnd_1_op','rnd_2_op','rnd_3_op','rnd_4_op','rnd_5_op','rnd_6_op','rnd_7_op')
  colnames(df_one)=col_names_1
  
  #now splitting each combined_rank columnn to get pre rating
  
  library(stringr)
  
  df_two$comb_rank=str_extract(df_two$comb_rank, "(?<=R: )\\d+")
  df_two$comb_rank=as.numeric(df_two$comb_rank)
  col_names_2=c('state','pre_rank','idk','col_rnd_1','col_rnd_2','col_rnd_3','col_rnd_4','col_rnd_5','col_rnd_6','col_rnd_7')
  colnames(df_two)=col_names_2
  head(df_two)
  
  
  
  #Now lets combine both data frames
  #first will remove columns we dont care about
  df_two$col_rnd_1=NULL
  df_two$col_rnd_2=NULL
  df_two$col_rnd_3=NULL
  df_two$col_rnd_4=NULL
  df_two$col_rnd_5=NULL
  df_two$col_rnd_6=NULL
  df_two$col_rnd_7=NULL
  df_two$idk=NULL
  head(df_two)
  
  #combine the two data frames
  
  chess_df=cbind(df_one,df_two)
  chess_df$player_id=as.integer(chess_df$player_id)
  head(chess_df)
  
  #creating new column named avg op rank
  #making a loop that will get all the oposition ranks
  
  
  chess_df$avg_op_pre_rank = NA
  
  # Loop through each row in chess_df
  for (i in 1:nrow(chess_df)) {
    
    # extract op ID for each round
    
    op_ids = unlist(chess_df[i, grep("rnd_\\d+_op", names(chess_df))])
    
    op_ids=as.integer(op_ids)
    
    #take out NA
    
    op_ids=na.omit(op_ids)
    
    # get prerank for each op
    
    op_ranks = chess_df[chess_df$player_id %in% op_ids, "pre_rank"]
    
    #mean op pre rank
    
    chess_df$avg_op_pre_rank[i] = mean(op_ranks, na.rm = TRUE)
  }
  
  head(chess_df)
  
  #making new dataframe with required format like the assignment states
  
  to_be_exported_list=list(Player_Name=chess_df$name,
                           Player_State=chess_df$state,
                           Total_Points=chess_df$total_pts,
                           Player_Pre_Rating=chess_df$pre_rank,
                           Average_Opponent_Pre_Rating=chess_df$avg_op_pre_rank)
  to_be_exported_df=as.data.frame(to_be_exported_list)
  
  head(to_be_exported_df)
  
  #write and export as csv
  write.csv(to_be_exported_df,export_path_name, row.names=FALSE)
  
}
```

```{r}
chess_to_csv("https://raw.githubusercontent.com/sleepysloth12/data607_proj1/main/tournamentinfo.txt", "C:/Users/bleac/OneDrive/Documents/Masters/Data607/week4/Project1/function_test.csv")
```

## Conclusion

In this project, I took the text file, processed it, and returned a csv file for the specifications asked for. I packaged all of that into a function that can be replicated. That being said, the function is not perfect. It will not work if a text file is inserted if it has a slightly different format. In the future, I can fix my function to account for this. Also, this assignment reminded me of object-oriented programming in python. One idea I had but didn't explore/ another way to do this project is by making a "Player" class, with each of the columns being an attribute; and creating a function that will extract the text from the file and assign them.
